Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-22 paper2.qmd
section:4; process:write,figures

<!-- 3. But are our specific problems? -->
However, Thurstone designed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precise trait measurement. He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270] because it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases). Therefore, it is surprising that, despite these limitations, CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp. 677].

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its limitations, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp. 16; @McElreath_2022]. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

@@

<!-- 4. How these judgment discrepancies can happen? -->
These discrepancies may arise from two possible causes. First, the discriminal processes of misfit texts may substantially overlap with those of non-misfit texts. Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale. Regarding the first cause, the overlap can occur when misfit texts share the same trait space and have equal dispersion as non-misfit texts or when they exhibit larger discriminal dispersions. Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second. @fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies. In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, where judges struggle to determine which text possesses more of the trait.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Thus, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.

@@

Finally, as discussed in the previous section, removing judges based on misfit statistics risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis. Researchers define *misfit judges* as those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp. 164-165; @Pollitt_2012b, pp. 289-290; @vanDaal_et_al_2016, pp. 4; @Goossens_et_al_2018, pp. 20].

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp. 269]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, showing how neglecting measurement error in an outcome can reduce the precision of a predictor' estimated effect.

@@

<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 5. So, how do SCMs and DAGs support identification? -->
Luckily, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in an interactive, non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

@@

<!-- 7. How the section is divided? -->
Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*. This model assumes that researchers have access to a "conceptual population" of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*. This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.

@@

@@

<!-- 2. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence. The absence of an edge implies no direct relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. They also display conditioned variables by surrounding them with a square. Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp. 32]. Although DAGs often omit exogenous variables for simplicity, this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 1. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli). The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.

@@

<!-- 2. the preliminary DAG for CJ -->
Moreover, @fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal outlined in Thurstone's theory. The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$. It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference. Additionally, the DAG differentiates between observed endogenous variables, such as the outcome, and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases.

@@

### The conceptual-population data structure {#sec-theory-theoretical_P2}

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system outlined in the previous section into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

$$
I_{n_{I}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J_{n_{J}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat12}

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$, where $n_{K}$ indicates the number of judgments each judge makes. Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns.

Additionally, we define the matrix $RPUV$ that assigns each row from the $IA$ matrix to a corresponding row from the $JK$ matrix. Thus, the matrix consists of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of repeated judgments each judge makes for every pair of stimuli generated by each pair of individuals in the population. The matrix is defined as follows:

@@

$$
RPUV_{n \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat3}

It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$. In this simple case, the vectors and matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:

$$
I_{5} = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
J_{3} = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$

<!-- 3. how de we use that data structure? -->
Then, using equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors capture all the observations from the conceptual population. Specifically, $O^{cp}_{RPUV}$ is an observed vector of length $n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, and $D_{RPUV}$ is a latent vector of the same dimension. Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
$$

@@

@@

>>>>

<!-- 3. But are our specific problems? -->
However, Thurstone designed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precise trait measurement. He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases). Therefore, it is surprising that, despite these limitations, CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp. 677].

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp. 16; @McElreath_2021]. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- the scale of the latent variable is arbitrary, so there is no loss of generality [@McElreath_2021] 02:40:30 - 02:40:45  -->

@@

<!-- 4. How these judgment discrepancies can happen? -->
These discrepancies may arise from two possible causes. First, the discriminal processes of misfit texts may substantially overlap with those of non-misfit texts. Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale. Regarding the first cause, the overlap can occur when misfit texts have equal dispersion and share the same trait space as non-misfit texts or when they exhibit larger discriminal dispersions. Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second. @fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies. In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, a space where judges struggle to determine which text possesses more of the trait.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
When the assumption of equal dispersions between stimuli does not hold, significant statistical and measurement issues can arise Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.

@@

Finally, as discussed in the previous section, removing judges based on misfit statistics risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis. *Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp. 164-165; @Pollitt_2012b, pp. 289-290; @vanDaal_et_al_2016, pp. 4; @Goossens_et_al_2018, pp. 20].

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp. 269]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.

@@

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear in the rest of the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 5. So, how do SCMs and DAGs support identification? -->
Luckily, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

@@

<!-- 7. How the section is divided? -->
Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*. This model assumes that researchers have access to a *conceptual population* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*. This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.

@@

<!-- 1. What is the assumption on the conceptual-population model -->
In the conceptual-population model, we assume an idealized scenario where researchers have access to a *conceptual population* of comparative data --data representing all repeated judgments made by each judge for every stimulus pair produced by each pair of individuals in the population. This assumption allows us to integrate Thurstone's theoretical principles and propose innovations to address some of the issues raised in @sec-theory-issues.

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli). The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.

@@

<!-- 3. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence. The absence of an edge implies no direct relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. They also display conditioned variables by surrounding them with a square. Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp. 32]. Although DAGs often omit exogenous variables for simplicity, this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 4. the preliminary DAG for CJ -->
@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal relationships outlined in Thurstone's theory. The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$. It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference. Additionally, the DAG differentiates between observed endogenous variables, such as the outcome (solid black circle), and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases (open circles).

@@

### The *conceptual population* data structure {#sec-theory-theoretical_P2}

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

$$
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat11}

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$. Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.

Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix. Thus, this matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population. Whe define the matrix as follows:

@@

$$
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat12}

It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$. In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:

$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$

<!-- 3. how de we use that data structure? -->
Now, using equations ([-@eq-mat11]) and ([-@eq-mat12]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{R}$, the texts' discriminal difference $D_{R}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors capture all the observations from the conceptual population. Specifically, $O^{cp}_{R}$ and $D_{R}$ are observed and latent vectors of length $n$, respectively. Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
$$

@@

<!-- 1. What is the more realistic scenario: the sample mechanism -->
In the sample-comparison model, we assume a more realistic scenario than the conceptual-population model. Specifically, we assume researchers have access to a data sample consisting of a limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, drawn from the conceptual population. Additionally, we assume that judges do not make *all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.

@@

<!-- Brought Rubin's missing data framework to the field. -->
